Applied Measure Theory for Probabilistic Modeling

### 2022-05-11 paper/paper.tex
edits

% a bit of a stupid exercise:
\newunicodechar{σ}{\ensuremath{\sigma}} %(U+03C3)
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{β}{\ensuremath{\beta}}
\newunicodechar{θ}{\ensuremath{\theta}}
\newunicodechar{μ}{\ensuremath{\mu}}
\newunicodechar{ν}{\ensuremath{\nu}}
\newunicodechar{⊙}{\ensuremath{\odot}}
\newunicodechar{κ}{\ensuremath{\kappa}}
\newunicodechar{η}{\ensuremath{\eta}}
\newunicodechar{∫}{\ensuremath{\int}}
\newunicodechar{ℓ}{\ensuremath{\ell}}
\newunicodechar{√}{\ensuremath{\sqrt{}}}
\newunicodechar{⊗}{\ensuremath{\otimes}}
\usepackage{booktabs}
\usepackage{multirow}
\renewcommand{\arraystretch}{1.2}

@@

Density computations are typically done in log space. For example, if we're interested in $\frac{\dd \mu}{\dd \nu}$, we'll typically instead compute \verb|logdensity(μ, ν, x)|. Here and below, we'll often write the math in terms of densities and the code in terms of log-densities, despite unfortunate inconsistency between the two.

@@

Working locally and in log-space also gives us a convenient \emph{antisymmetry},
\begin{verbatim}
logdensity(μ, ν, x) == -logdensity(ν, μ, x)
\end{verbatim}

In particular, \verb|logdensity(μ, ν, x)| takes special values in some common cases:
\newcommand{\notllx}{\mathrel{\centernot{\ll}\!\!_x}}

@@

\section{Base measures and density recursion}
To this point, we've described the implementation in terms of the three-argument \verb|logdensity(μ, ν, x)|. But implementing things directly in these terms would require a definition for every pair of possible measures, which is of course intractable.

We address this by instead assigning every measure a \emph{base measure}. This is defined locally, with default method
\begin{verbatim}
basemeasure(μ::AbstractMeasure, x) = basemeasure(μ)
\end{verbatim}
This allows users to define a base measure specific to a given neighborhood if they prefer, or to define a global base measure when that suffices.

The log-density is then fundamentally expressed in term of the two-argument \verb|logdensity(μ, x)|, which gives the log-density with respect to this base measure. Thus we have
\begin{verbatim}
logdensity(μ, x) == logdensity(μ, basemeasure(μ, x), x)
\end{verbatim}

>>>>

% a bit of a stupid exercise:
\newunicodechar{σ}{\ensuremath{\sigma}} %(U+03C3)
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{β}{\ensuremath{\beta}}
\newunicodechar{θ}{\ensuremath{\theta}}
\newunicodechar{μ}{\ensuremath{\mu}}
\newunicodechar{ν}{\ensuremath{\nu}}
\newunicodechar{⊙}{\ensuremath{\odot}}
\newunicodechar{κ}{\ensuremath{\kappa}}
\newunicodechar{η}{\ensuremath{\eta}}
\newunicodechar{∫}{\ensuremath{\int}}
\newunicodechar{ℓ}{\ensuremath{\ell}}
\newunicodechar{√}{\ensuremath{\sqrt{}}}
\newunicodechar{⊗}{\ensuremath{\otimes}}
\newunicodechar{π}{\ensuremath{\pi}}
\newunicodechar{ℝ}{\ensuremath{\mathbb{R}}}
\usepackage{booktabs}
\usepackage{multirow}
\renewcommand{\arraystretch}{1.2}

@@

Density computations are typically done in log space. For example, if we're interested in $\frac{\dd \mu}{\dd \nu}$, we'll typically instead compute \verb|logdensity_rel(μ, ν, x)|.\footnote{Density and log-density are \emph{always} relative; the \texttt{rel} here is to indicate that we want the density relative to a second measure which is explicitly specified.}
Here and below, we'll often write the math in terms of densities and the code in terms of log-densities, despite unfortunate inconsistency between the two.

@@

Working locally and in log-space also gives us a convenient \emph{antisymmetry},
\begin{verbatim}
logdensity_rel(μ, ν, x) == -logdensity_rel(ν, μ, x)
\end{verbatim}

In particular, \verb|logdensity_rel(μ, ν, x)| takes special values in some common cases:
\newcommand{\notllx}{\mathrel{\centernot{\ll}\!\!_x}}

@@

\section{Base measures and density recursion}
It's very useful for users to be able to \emph{call} \verb|logdensity_rel(μ, ν, x)|. But \emph{defining} things directly in these terms would require a definition for every pair of possible measures, which is of course intractable.

We'll now discuss the mechanics of defining a measure, followed by {\bf XXX description of the next subsection XXX}

\subsection*{Defining a measure}

To define a new measure requires a \emph{base measure}, the \emph{log-density} with respect to that base measure, and a \emph{support}. 
A base measure can be defined locally or globally. In the latter case, the local base measure defaults to 
\begin{verbatim}
basemeasure(μ::AbstractMeasure, x) = basemeasure(μ)
\end{verbatim}
This allows users to define a base measure specific to a given neighborhood if they prefer, or to define a global base measure when that suffices.

For example, for a standard \verb|Normal| measure, we have

\begin{verbatim}
basemeasure(::Normal{()}) = (1/sqrt2π) * Lebesgue(ℝ)
logdensity_def(::Normal{()} , x) = - x^2 / 2 
insupport(::Normal{()}, x) = true
\end{verbatim}

Note the difference from a typical distribution-oriented implementation; we have changed the base measure in order to simplify the log-density. This leads to better compositional efficiency.
